---
title: Trucks! Trucks! Trucks!
author: Blake Shaffer
date: '2018-12-17'
slug: trucks-trucks-trucks
categories: []
tags: []
header:
  caption: ''
  image: ''
twitterImg: post/2018-12-17-trucks-trucks-trucks_files/figure-html/truckplot-1.png
---
Quick post to show how easy it is to do quick data exploration using R. This post walks us through making a (nice!) chart of **new vehicle sales in Canada**, breaking it down by vehicle type and region.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, cache=TRUE)
```

## Step 1: Load the three packages we will use for this plot:

* `tidyverse` : because always tidyverse
* `cansim` : great package to easily extract Statistics Canada data
* `ggthemes` : not necessary, but a quick way to pretty up your ggplot


```{r}
library(tidyverse)
library(cansim)
library(ggthemes)
```

## Step 2: Data wranglin'

1. Pull in the data
2. Filter to what we want in the data
3. Fiddle with some labels

```{r}
data <- get_cansim(20100001) %>% normalize_cansim_values()

plotdata <- data %>%
  filter(GEO!="Canada",
         Sales=="Units",
         `Vehicle type` %in% c("Passenger cars","Trucks"),
         `Origin of manufacture`=="Total, country of manufacture")

# Change some GEO labels to shorten them
plotdata$GEO[plotdata$GEO=="Newfoundland and Labrador"] <- "Nfld & Lab."
plotdata$GEO[plotdata$GEO=="British Columbia and the Territories"] <- "BC & Terr."
plotdata$GEO[plotdata$GEO=="Prince Edward Island"] <- "PEI"
plotdata$GEO <- factor(plotdata$GEO, levels=c("BC & Terr.","Alberta","Saskatchewan","Manitoba","Ontario","Quebec","New Brunswick","PEI","Nova Scotia","Nfld & Lab."))

# Change "passenger cars" to "cars" for plot label
plotdata$`Vehicle type`[plotdata$`Vehicle type`=="Passenger cars"] <- "Cars"
```

## Step 3: Plot it!

```{r truckplot, fig.width=9, fig.height=5}
ggplot(plotdata, aes(Date,VALUE,color=`Vehicle type`))+
  geom_point(size=.1,alpha=.2)+
  geom_smooth(se=F)+
  facet_wrap(~GEO, scales="free_y", nrow=2)+
  scale_color_ptol(name="")+  # nice color scheme from ggthemes package
  scale_x_date(limits=c(as.Date("1980-01-01"),as.Date("2020-01-01")),
               breaks = c(as.Date("1980-01-01"),as.Date("2000-01-01"),as.Date("2020-01-01")),
               date_labels = "%Y")+
  theme_tufte(12,"Shree Devanagari 714")+     # font similar to 'Avenir'
  theme(legend.position = "top",
        title=element_text(face="bold"),
        plot.caption = element_text(face="italic"),
        legend.text = element_text(size=14))+
  labs(x="", y="",
       subtitle="New vehicle sales (monthly)",
       title="Trucks! Trucks! Trucks!",
       caption="Fitted line through monthly (seasonally unadjusted) data.\nSource: Statscan NDM Table 20-10-0001.")
```

***

There you go. A nice quick plot showing a pretty stark shift in consumer preferences towards trucks across *all regions* in Canada. An interesting extension of this data exploration would be to see what this means in terms of fleet-wide fuel economy. Are people switching from Civics to F-150s? Or Crown Vics to Subaru Crosstreks? The latter would mean a shift from cars to "trucks" is improving overall gas mileage... More data needed!

*Happy Charting!*

*-Blake*