---
title: 'Gasoline demand: BC and the Rest of Canada'
author: Blake Shaffer
date: '2018-11-30'
slug: gasoline-demand-bc-and-the-rest-of-canada
categories:
  - energy
tags: []
twitterImg: post/2018-11-30-gasoline-demand-bc-and-the-rest-of-canada_files/figure-html/gas_summary-1.png
header:
  caption: ''
  image: ''
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, cache=TRUE)
```

This post looks at gasoline demand across the Canadian provinces and territories. It was prompted by this tweet by @sjmuir, noting that gasoline demand per capita in BC is unchanged since 2008, the year of implementation of the BC carbon tax.

<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">It&#39;s taboo to point this out, but the fact is that British Columbia residents used as much gasoline in 2017 as they did in 2008. That&#39;s why I&#39;m skeptical whenever it&#39;s stated that carbon pricing has led to some kind of revolution in fuel usage, because it hasn&#39;t. <a href="https://t.co/3IPILHxnTw">pic.twitter.com/3IPILHxnTw</a></p>&mdash; Stewart Muir (@sjmuir) <a href="https://twitter.com/sjmuir/status/1068523879331586048?ref_src=twsrc%5Etfw">November 30, 2018</a></blockquote>
<script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

The data didn't fit my prior as I had seen in a [recent paper](https://papers.ssrn.com/sol3/papers.cfm?abstract_id=2778868) that shows BC gasoline demand per capita falling relative to the rest of Canada.

So what gives?

It turns out, a lot has changed since the above referenced paper. Namely, BC gasoline demand has risen quite a bit since 2015. Let's dig in. I'll first plot the change in total gasoline demand in each province since 2008:

```{r gasoline_index, fig.width=8,fig.height=5}
library(tidyverse)
library(lubridate)
library(cansim)
library(ggrepel)
library(ggthemes)

## GET EMPLOYMENT DATA
data.gas <- get_cansim(23100066) %>% 
  normalize_cansim_values() %>%
  filter(`Type of fuel sales`=="Net sales of gasoline",
         GEO!="Northwest Territories including Nunavut") %>%
  mutate(Year=as.numeric(REF_DATE),
         Region = case_when(GEO %in% c("British Columbia","Alberta","Saskatchewan") ~ "West",
                            GEO %in% c("Manitoba","Ontario","Quebec")  ~ "Central",
                            GEO %in% c("New Brunswick","Nova Scotia","Prince Edward Island","Newfoundland and Labrador") ~ "East",
                            GEO %in% c("Yukon","Northwest Territories","Nunavut") ~ "Territories",
                            GEO=="Canada" ~ "CANADA"),
         GEO.abb = case_when(GEO=="British Columbia" ~ "BC",
                             GEO=="Alberta" ~ "AB",
                             GEO=="Saskatchewan" ~ "SK",
                             GEO=="Manitoba" ~ "MB",
                             GEO=="Ontario" ~ "ON",
                             GEO=="Quebec" ~ "QC",
                             GEO=="New Brunswick" ~ "NB",
                             GEO=="Nova Scotia" ~ "NS",
                             GEO=="Prince Edward Island" ~ "PE",
                             GEO=="Newfoundland and Labrador" ~ "NL",
                             GEO=="Yukon" ~ "YT",
                             GEO=="Northwest Territories" ~ "NT",
                             GEO=="Nunavut" ~ "NU",
                             GEO=="Canada" ~ "CAN")) %>%
  group_by(GEO) %>%
  mutate(Index=VALUE/VALUE[Year==2008]) %>%
  dplyr::select(Year,Date,GEO,GEO.abb,Region,Index,"Gasoline"=VALUE)
data.gas$Region <- factor(data.gas$Region, levels=c("CANADA","West","Central","East","Territories"))

ggplot()+
  geom_line(data=data.gas %>% filter(GEO!="Canada",GEO!="Nunavut",GEO!="British Columbia"), 
            aes(Year,Index,group=GEO,color=GEO),
            size=1)+
  geom_line(data=data.gas %>% filter(GEO=="British Columbia"),
            aes(Year,Index),
            size=2, color="firebrick")+
  geom_label_repel(data=data.gas %>% filter(Year==2017,GEO!="Canada",GEO!="Nunavut",GEO!="British Columbia"),
                   aes(Year, Index, color=GEO, label=GEO.abb),
                   nudge_x=.5)+
  geom_label_repel(data=data.gas %>% filter(Year==2017,GEO=="British Columbia"), 
                   aes(Year, Index, label=GEO.abb),
                   nudge_x=.5, color="firebrick")+
  geom_hline(yintercept = 1,linetype="dashed")+
  scale_y_continuous(labels=scales::percent)+
  scale_x_continuous(breaks=seq(2008,2018,2),limits=c(2008,2018))+
  coord_cartesian(ylim=c(.95,1.4))+
  scale_color_viridis_d(name="")+
  theme_classic(14,"Gill Sans")+
  theme(plot.caption=element_text(size=8,face="italic"),
        legend.position = "none")+
  labs(x="",y="",
       subtitle="Gasoline demand relative to 2008 (unadjusted for population)",
       caption="Statistics Canada NDM Table 23-10-0066\nChart by @bcshaffer")
```

It's a busy chart but there's a few things to note:

  1. Sask is guzzling the stuff. This isn't per capita yet, but still!
  2. Meanwhile, BC, which showed a big dip in 2011-2014 has come roaring back.
  
The above chart isn't per capita. So let's get quarterly population estimates, annualize them, and create a gasoline_per_capita variable:

```{r per_capita, fig.width=10,fig.height=4}
## GET POPULATION DATA
data.population <- get_cansim(0510005) %>% 
  normalize_cansim_values() %>%
  mutate(Year=year(Date)) %>%
  dplyr::select(Year,GEO,"Population"=VALUE) %>%
  group_by(GEO,Year) %>%
  summarise(Population=mean(Population)) 

## Merge and plot
plotdata <- left_join(data.gas,data.population,by=c("Year","GEO"))
plotdata$gas_per_cap <- plotdata$Gasoline/plotdata$Population/365
ggplot(plotdata, aes(Year,gas_per_cap,group=GEO,color=GEO,label=GEO.abb))+
  geom_line(size=1.5)+
  geom_vline(xintercept = 2008,linetype="dashed")+
  geom_label_repel(data=plotdata %>% filter(Year==2017),nudge_x=.5)+
  facet_wrap(~Region,nrow=1)+
  scale_x_continuous(breaks=seq(2008,2018,4),limits=c(2008,2019))+
  scale_color_viridis_d(name="",option="B")+
  theme_classic(14,"Gill Sans")+
  theme(plot.caption=element_text(size=8,face="italic"),
        legend.position = "none")+
  labs(x="",y="Litres per day",
       subtitle="Per capita gasoline demand",
       caption="Statistics Canada NDM Table 23-10-0066 and 17-10-0009\nChart by @bcshaffer")
```


Another question, is how different economic growth--rather than just population growth--across provinces impacts their gasoline demand. So here we look at *gasoline intensity*, the ratio of annual gasoline demand (in litres) per thousand dollars of GDP. (Note: I drop the territories as their scale gets wonky. Sorry Territories...)

```{r gasoline_intensity, fig.width=10,fig.height=4}
## GET GDP DATA
data.GDP <- get_cansim(36100222) %>% 
  normalize_cansim_values() %>%
  filter(Prices=="Chained (2012) dollars",
         Estimates=="Gross domestic product at market prices") %>%
  mutate(Year=year(Date)) %>%
  dplyr::select(Year,GEO,"GDP"=VALUE) 

plotdata <- left_join(plotdata,data.GDP, by=c("Year","GEO"))
plotdata$gas_intensity <- plotdata$Gasoline / plotdata$GDP * 1000
plotdata <- plotdata %>% filter(Region!="Territories")

ggplot(plotdata, aes(Year,gas_intensity,group=GEO,color=GEO,label=GEO.abb))+
  geom_line(size=1.5)+
  geom_vline(xintercept = 2008,linetype="dashed")+
  geom_label_repel(data=plotdata %>% filter(Year==2017),nudge_x=1)+
  facet_wrap(~Region,nrow=1)+
  scale_x_continuous(breaks=seq(1995,2015,10),limits=c(1994,2020))+
  coord_cartesian(ylim=c(16,45))+
  scale_color_viridis_d(name="",option="B")+
  theme_classic(14,"Gill Sans")+
  theme(plot.caption=element_text(size=8,face="italic"),
        legend.position = "none")+
  labs(x="",y="L per $1000 GDP",
       subtitle="Gasoline intensity\n(Gas demand / GDP)",
       caption="Statistics Canada NDM Table 23-10-0066 and 36-10-0222\nChart by @bcshaffer")
```

It's a little hard to tell in that graph, but in the **West** plot, BC is the line that starts highest and is falling rapidly throughout the period. I'll re-plot it on one panel here:

```{r gas_intensity_one_chart, fig.width=6, fig.height=5}
ggplot()+
  geom_line(data=plotdata %>% filter(GEO!="Nunavut",GEO!="British Columbia"), 
            aes(Year,gas_intensity,group=GEO,color=GEO),
            size=1)+
  geom_line(data=plotdata %>% filter(GEO=="British Columbia"),
            aes(Year,gas_intensity),
            size=2, color="firebrick")+
  geom_label_repel(data=plotdata %>% filter(Year==2017,GEO!="Nunavut",GEO!="British Columbia"),
                   aes(Year, gas_intensity, color=GEO, label=GEO.abb),
                   nudge_x=1)+
  geom_label_repel(data=plotdata %>% filter(Year==2017,GEO=="British Columbia"), 
                   aes(Year, gas_intensity, label=GEO.abb),
                   nudge_x=.5, color="firebrick")+
  geom_vline(xintercept = 2008,linetype="dashed")+
  scale_x_continuous(breaks=seq(1995,2015,10),limits=c(1994,2020))+
  coord_cartesian(ylim=c(16,48))+
  scale_color_viridis_d(name="")+
  theme_classic(14,"Gill Sans")+
  theme(plot.caption=element_text(size=8,face="italic"),
        legend.position = "none")+
  labs(x="",y="L per $1000 GDP",
       subtitle="Gasoline intensity\n(Gas demand / GDP)",
       caption="Statistics Canada NDM Table 23-10-0066 and 36-10-0222\nChart by @bcshaffer")
```


To summarise, and make it clearer, let's take a look at just BC and Canada side by side for both gasoline intensity and gasoline per capita:

```{r gas_summary, fig.width=8,fig.height=5}
data.summary <- plotdata %>%
  filter(GEO %in% c("British Columbia","Canada")) %>%
  dplyr::select(Year,GEO,GEO.abb,gas_per_cap,gas_intensity) %>%
  gather(key,value,gas_per_cap,gas_intensity)

ggplot(data.summary, aes(Year,value,group=GEO,color=GEO))+
  geom_line(size=1.5)+
  geom_vline(xintercept = 2008,linetype="dashed")+
  facet_wrap(~key,scales="free_y")+
  scale_x_continuous(breaks=seq(1995,2015,5))+
  scale_color_manual(name="",values=c("forestgreen","red"))+
  theme_classic(12,"Gill Sans")+
  theme(plot.caption=element_text(size=8,face="italic"),
        legend.position = "top",
        legend.text = element_text(size=14),
        strip.text = element_text(size=12))+
  labs(x="",y="",
       subtitle="gas_intensity: L/GDP ($'000s)\ngas_per_cap: L/capita/day",
       title="BC vs the Rest of Canada",
       caption="Statistics Canada NDM Table 23-10-0066, 36-10-0222, 17-10-0009\nChart by @bcshaffer")
```

**Bottom line: gasoline intensity (the amount of gasoline demand per $1000 of GDP) is falling across the board. That's a good thing. B.C.'s gasoline intensity is falling more rapidly than the Canadian average. On a per capita basis, despite rising in the past 2 years on the back of strong GDP growth, BC is still falling relative to the Canadian average.**